---
title: "Clustering"
output: pdf_document
layout: page
---

### K-means

We'll use the tissues gene expression file to look at clustering. There are expressions for 100 genes in 1816 samples. The samples come from 5 different tissues.

```{r}
library(rafalib)
e <- read.delim("data/expr.txt", row.names=1)
tab <- read.delim("data/class_labels.txt", row.names=1)
head(tab) # rows represent samples
head(e)[,1:2] # rows represent genes
```

We train the `kmeans` model in the **`stats`** package on the first 5 samples for total cluster number = 5, for a maximum of 10 iterations.

```{r, echo=FALSE}
tissue <- tab[["tissue"]]
```
```{r}
set.seed(1)
km <- kmeans(t(e), centers=t(e)[1:5,], iter.max=10)
cbind(cluster=c(1,2,3,4,5), samples=km$size)
table(tissue=tissue, cluster=km$cluster)
```

### Bayesian Information Criterion

We'll now optimize the number of clusters (k) using Bayesian Information Criterion,
$$BIC = -2 \log \hat{L} + m \log n$$

```{r}
bic <- function(fit) {
  k <- nrow(fit$centers) # number of clusters
  m <- ncol(fit$centers) * k # number of free parameters
  n <- length(fit$cluster) # number of samples
  D <- fit$tot.withinss
  return (D + m*log(n))
}
```

In a k-means model, the number of free parameters *m* is just the number of clusters *k* times *d*-dimesional points. In our case, *d* = 100 (the number of genes), and *k* = 5 (by choice), so **_m_ = 500**.

Now, let us compute the BIC scores for a k-means model trained using *k* = 2 to 10. In each case, the initial centers are the first *k* points.

```{r}
bic.values <- c()
for (k in 2:10) {
  bic.values[k-1] <- bic(kmeans(t(e), centers=t(e)[1:k,], iter.max=10))
}
par(mfrow = c(1, 1))
plot(c(2:10), bic.values, xlab="Number of clusters (k)", ylab="BIC scores", pch=1)
data.frame(k=c(2:10), bic=bic.values)
```

We see that the BIC score is minimum for **k = 5**. Thus, this is the optimal number of clusters for our example.

### Visualizing our output

```{r, fig.height=5, fig.align='center'}
d <- dist(t(e)) # distance between sample points
km <- kmeans(t(e), centers=5)
mds <- cmdscale(d)

mypar(1,1)
plot(mds[,1], mds[,2], col=km$cluster, pch=16)
legend("bottomleft", legend=c(1:5), col=c(1:5), pch=16)
```
