---
title: "Principal Components Analysis"
output: pdf_document
layout: page
---

### Pre-processing data

We'll be using the gene expression dataset for 17580 genes from 73 samples. There are two phenotypes, 0:no-diesase and 1:Parkinson's. We have an additional dataset containing 3 sample covariates.

```{r}
library(rafalib)

e <- read.delim("data/counts.txt", row.names=1)
tab <- read.delim("data/phen.txt", row.names=1)
c <- read.delim("data/cov.txt", row.names=1)
```

We take the log-transform (normalize) of gene expression data and calculate the Z-score (standardize).

```{r}
e_prime <- t(e) # Re-order data
L <- log2(1 + e_prime) # Log-transform data
head(L)[,1:2]
Z <- scale(L) # Z-score
head(Z)[,1:2]
```

### Computing principal components and percent variance

We use the `prcomp` function in the **`stats`** package to compute PCs of the scaled data.

```{r}
pca <- prcomp(Z)
pca$sdev[1:10]
pca$rotation[1:5,1:2]
```

We now extract the variances of the components, and report the percent variance.

```{r}
pca.var <- pca$sdev^2
pca.percent_var <- pca.var/sum(pca.var)
pca.percent_var[1:10]
```

### Scatter plot of first two PC loadings

```{r}
cols <- c('pink', 'red')
par(mfrow = c(1, 1))
plot(pca$x[,1], pca$x[,2], col=cols[tab$disease+1], pch=16, xlab="PC1", ylab="PC2")
legend("bottomleft", legend=c("Normal","Parkinson's"), col=cols, pch=16)
```

### Pairwise Pearson correlation between PCs and covariates

Now, we want to see if any of the top 10 PCs are strongly correlated with any of the given covariates. Let's begin by looking at all the correlation estimates.

```{r}
c_prime <- data.frame(t(c))
cor(pca$x[,1:10], c_prime, use="pairwise.complete.obs", method="pearson")
```

Let us say a PC is strongly correlated to a covariate if correlation estimate |*r*| > 0.2, and p-value < 0.05. Then, PCs strongly correlated to *post mortem interval* are

```{r}
for (i in 1:10) { # For top 10 PCs
  c_test <- cor.test(pca$x[,i], c_prime[,1], use="pairwise.complete.obs", method="pearson")
  if (abs(c_test$estimate) > 0.2 & c_test$p.value < 0.05) {
    cat("PC", i, sep="")
    cat(" estimate =", c_test$estimate, "p-value =", c_test$p.value, "\n", sep=" ")
  }
}
```

Similarly, those strongly correlated to *RNA integrity number* are

```{r echo=FALSE}
for (i in 1:10) { # For top 10 PCs
  c_test <- cor.test(pca$x[,i], c_prime[,2], use="pairwise.complete.obs", method="pearson")
  if (abs(c_test$estimate) > 0.2 & c_test$p.value < 0.05) {
    cat("PC", i, sep="")
    cat(" estimate =", c_test$estimate, "p-value =", c_test$p.value, "\n", sep=" ")
  }
}
```

and those to *age*

```{r echo=FALSE}
for (i in 1:10) { # For top 10 PCs
  c_test <- cor.test(pca$x[,i], c_prime[,3], use="pairwise.complete.obs", method="pearson")
  if (abs(c_test$estimate) > 0.2 & c_test$p.value < 0.05) {
    cat("PC", i, sep="")
    cat(" estimate =", c_test$estimate, "p-value =", c_test$p.value, "\n", sep=" ")
  }
}
```

### Pairwise Pearson correlation between disease status and covariates

```{r}

```