---
title: "Principal Components Analysis"
output: pdf_document
layout: page
---

### Pre-processing data

We'll be using the gene expression dataset for 17580 genes from 73 samples. There are two phenotypes, 0:no-diesase and 1:Parkinson's. We have an additional dataset containing 3 sample covariates.

```{r}
library(rafalib)

e <- read.delim("data/counts.txt", row.names=1)
tab <- read.delim("data/phen.txt")
c <- read.delim("data/cov.txt", row.names=1)
```

We take the log-transform (normalize) of gene expression data and calculate the Z-score (standardize).

```{r}
e_prime <- t(e) # Re-order data
L <- log2(1 + e_prime) # Log-transform data
head(L)[,1:2]
Z <- scale(L) # Z-score
head(Z)[,1:2]
```

### Computing principal components and percent variance

We use the `prcomp` function in the **`stats`** package to compute PCs of the scaled data.

```{r}
pca <- prcomp(Z)
pca$sdev[1:10]
pca$rotation[1:5,1:2]
```

We now extract the variances of the components, and report the percent variance.

```{r}
pca.var <- pca$sdev^2
pca.percent_var <- pca.var/sum(pca.var)
pca.percent_var[1:10]
```

### Plots of first two PC loadings

```{r}
cols <- c('pink', 'red')
par(mfrow = c(1, 1))
plot(pca$x[, 1], pca$x[, 2], col=cols[tab$disease+1], pch=16, xlab="PC1", ylab="PC2")
legend("bottomleft", legend=c("Normal","Parkinson's"), col=cols, pch=16)
```
